---
title: "topologyR: User Manual"
subtitle: "Topologies of graphs, bitopologies of time series and visibility graphs (version 0.4.0)"
author: "José Mauricio Gómez Julián"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = "#>")
```

```{r load}
library(topologyR)
```

# Overview

`topologyR` turns a graph into a topology on its vertices and computes what can
be said about that topology exactly. The topology is the one generated by the
*closed neighbourhoods* of the vertices: each vertex together with its
neighbours. For time series, the graph is a *visibility graph*, horizontal or
natural, built at equally or unequally spaced instants; its directed version,
with arcs from the earlier to the later observation, yields two topologies, a
forward one and a backward one, which together form a bitopological space. The
package computes connected components exactly without enumerating any
topology, enumerates topologies when they are small enough, and reports
invariants of the forward-backward pair.

Three results organize the package and are proved in this manual:

1. In a finite space, the connected components are the classes of the
   equivalence generated by comparability in the specialization preorder, and
   that preorder can be read off the subbase. The components are therefore
   exact even when a computation is truncated (the section on connectedness and connected components).
2. The forward and backward bases have the same number of sets, for every
   digraph (the section on the equality of the two base sizes).
3. Reading a series backwards, with its instants reflected, exchanges the
   forward and backward topologies (theorem T1, the section on theorem T1).

The natural visibility criterion is decided exactly on the numbers the
function receives, under the hypotheses on the floating-point arithmetic
stated in the help page `?natural_visibility_exactness` (the section on the
natural visibility graph). What the package does not do is
equally important: none of its invariants is a test of the time
reversibility of a stochastic process (the section on what the invariants do not say).

# Foundations

## Topologies, bases and subbases

A *topology* on a set $X$ is a family $\tau$ of subsets of $X$, its *open
sets*, that contains $\emptyset$ and $X$ and is closed under arbitrary unions
and finite intersections. Complements of open sets are *closed*. A family
$\mathcal{B} \subseteq \tau$ is a *base* of $\tau$ when every open set is a
union of members of $\mathcal{B}$. A family $\mathcal{S}$ of subsets of $X$
*generates* the smallest topology that contains it; its open sets are the
unions of finite intersections of members of $\mathcal{S}$, the empty
intersection being $X$, and $\mathcal{S}$ is called a *subbase* of that
topology. The finite intersections of a subbase form a base.

The number of sets in the base produced from a subbase is a property of the
construction, not of the topology. On $\{1, 2, 3\}$, the subbases
$\{\{1\},\{2\},\{3\}\}$ and $\{\{1,2\},\{2,3\},\{1,3\}\}$ both generate the
discrete topology, in which every subset is open, but the first is its own
closure under intersections, with three sets, while the closure of the second
adds the three singletons and has six.

## Finite spaces and the specialization preorder

On a finite set, every family of open sets is finite, so the intersection of
all open sets containing a point $x$ is open: it is the smallest open set
containing $x$, written $U_x$. If the topology is generated by a subbase
$\mathcal{S}$, then $U_x = \bigcap \{ s \in \mathcal{S} : x \in s\}$, with
$U_x = X$ when no member contains $x$. A set is open if and only if it
contains $U_x$ for each of its points.

The *specialization preorder* is $x \le y \iff y \in U_x$: every open set that
contains $x$ contains $y$. It can be read off any family $F$ with
$\mathcal{S} \subseteq F \subseteq \tau$: every member of $F$ that contains $x$
contains $y$ if and only if $y \in U_x$. Indeed, if $y \in U_x$, every open set
containing $x$ contains $U_x$; conversely, the members of $\mathcal{S}$ that
contain $x$ are among those of $F$, and their intersection is $U_x$.

## Connectedness and connected components

A space is *connected* when it is not the union of two disjoint non-empty
open sets; equivalently, when its only subsets that are both open and closed
are $\emptyset$ and $X$. A *connected component* is a maximal connected
subset; the components partition the space.

**Proposition.** *In a finite space, the connected components are the classes
of the equivalence relation $\sim$ generated by comparability ($x$ and $y$ are
comparable when $x \le y$ or $y \le x$).*

*Proof.* Each class $C$ is open, because it contains $U_x$ for each of its
points ($y \in U_x$ means $x \le y$), and closed, because its complement is a
union of other classes; so every connected component lies inside one class.
Each class is connected: if $C$ were the union of two disjoint non-empty sets
$A$ and $B$, both open relative to $C$, some $x \in A$ and $y \in B$ would be
comparable, since $C$ is a single class; if $x \le y$, the open set that cuts
out $A$ contains $U_x$ and hence $y$, which is impossible, and the case
$y \le x$ is symmetric. $\square$

The proposition reduces connectedness to a graph search. Since the preorder
can be read off any family between the subbase and the topology, so can the
components, which is why `topologyR` computes them from a base that may have
been truncated (the section on the states of a computation). The proposition also shows what a
connected *graph* does and does not imply for the topology of its closed
neighbourhoods: every closed neighbourhood lies inside one component of the
graph, so a disconnected graph gives a disconnected topology, but a connected
graph can give a disconnected topology.

```{r path-graph}
path <- list(2L, c(1L, 3L), c(2L, 4L), 3L)
generate_topology(path, 4L, max_open_sets = 0L)$components
```

On the path $1 - 2 - 3 - 4$, the closed neighbourhoods $\{1,2\}$ and
$\{3,4\}$ are open and complementary, so the topology splits although the
graph is connected.

# From a graph to a topology

## Closed neighbourhoods

`generate_topology()` takes adjacency lists and uses as subbase the closed
neighbourhoods $N[v] = \{v\} \cup N(v)$. Nada, El Atik and Atef (2018,
Definition 3.1) take as subbase the post classes of the adjacency relation,
which for a loopless graph are the open neighbourhoods $N(v)$; `topologyR`
uses the closed variant, so that every vertex lies in its own subbase element,
the subbase covers the vertex set, and $U_v \subseteq N[v]$. The two
constructions give different topologies. On the complete graph every closed
neighbourhood is the whole vertex set, and the closed variant gives the
indiscrete topology:

```{r complete-graph}
complete_topology(1:6)[c("n_open_sets", "connected")]
```

whereas the open neighbourhoods $V \setminus \{v\}$ meet in every singleton
and give the discrete topology.

## The base, by wavefront intersection

The base is the closure of the subbase under non-empty finite intersections.
The engine computes it by wavefront propagation: each set added in a round is
intersected, in the next round, with every set already in the base, until a
round adds nothing. Every set added in a round is a proper subset of a set
added in the previous one, so the closure ends after at most $n$ rounds; the
base itself, however, can have a number of sets exponential in $n$, and the
argument `max_base_sets` stops the closure at the first new intersection that
makes the base exceed that number.

## Components without enumeration

The components are computed from the signatures
$S(x) = \{B \in \mathcal{B} : x \in B\}$, stored as bitsets: $x$ and $y$ are
comparable when $S(x) \subseteq S(y)$ or $S(y) \subseteq S(x)$, and a
breadth-first search over the comparability graph returns the classes. With
$b$ base sets this takes $O(n^2 \lceil b/64 \rceil)$ word operations. Because
the engine inserts the whole subbase before the closure can be cut, the
signatures always encode the specialization preorder, and the components are
exact.

## Enumerating the topology

The open sets are enumerated, only if `max_open_sets` is positive, by the
same wavefront scheme with unions. There can be up to $2^n$ of them, so the
enumeration is meant for small spaces or for checks that need the open sets
themselves. When the space is already known to be disconnected, the engine
warns that the enumeration is not needed for connectivity.

## The states of a computation

| Base | Enumeration | `base_complete` | `topology_complete` | Components |
|:-----|:------------|:----------------|:--------------------|:-----------|
| complete | not requested | `TRUE` | `NA` | exact |
| complete | complete | `TRUE` | `TRUE` | exact |
| complete | truncated | `TRUE` | `FALSE` | exact |
| truncated | not requested | `FALSE` | `NA` | exact |
| truncated | run | `FALSE` | `FALSE` | exact |

`topology_complete` is `TRUE` only when the enumerated family is the whole
topology. A truncated base contains the subbase but not all its
intersections, and the union closure of such a family is in general not the
topology:

```{r truncated}
g <- natural_visibility_graph(c(3, 1, 4, 1, 5, 9, 2, 6))
full <- suppressWarnings(generate_topology(g$adjacency, g$n, max_open_sets = 1e5))
cut <- suppressWarnings(generate_topology(g$adjacency, g$n, max_open_sets = 1e5,
                                          max_base_sets = 12L))
c(full = full$n_open_sets, truncated_base = cut$n_open_sets)
c(full$topology_complete, cut$topology_complete)
identical(full$components, cut$components)
```

(The warnings that the two calls suppress are the engine's notice that the
space is disconnected, so that its enumeration is not needed for
connectivity.)

# Visibility graphs

## The horizontal visibility graph

Two observations of a series are adjacent in the horizontal visibility graph
(Luque et al., 2009) when every observation strictly between them is strictly
lower than both. Consecutive observations are always adjacent. The criterion
compares values and nothing else, so the graph is computed exactly on the
numbers received, is invariant under any strictly increasing transformation
of the values, and does not depend on the spacing of the instants.

The engine scans the series once with a stack whose values decrease strictly
from bottom to top. When observation $i$ arrives, every stacked index $j$ with
$y_j \le y_i$ is popped and joined to $i$: nothing between them reaches $y_j$,
or it would have popped $j$ earlier. The index left on top, if any, is higher
than $y_i$ and is joined to $i$ unless a popped value equals $y_i$, which
blocks at height $\min = y_i$. Every other earlier index is blocked, either by
that top or by the index that popped it. The algorithm runs in $O(n)$ time.

## The natural visibility graph

Two observations $a < b$ are adjacent in the natural visibility graph
(Lacasa et al., 2008) when every observation $k$ between them lies strictly
below the segment joining $(t_a, y_a)$ to $(t_b, y_b)$, that is, when

$$D(a, b, k) = y_a (t_b - t_k) + y_b (t_k - t_a) - y_k (t_b - t_a) > 0 .$$

Every horizontal edge is a natural edge, because a point lower than both ends
of a segment is below it. Lacasa et al. (2008) describe the graph as
inheriting properties of the series: in the words of their abstract, periodic
series convert into regular graphs, random series into random graphs, and
fractal series into scale-free networks. The abstract does not say for which
series these properties hold, and they do not hold for every finite series:
the periodic series 0, 1, 0, 1, 0, at the equally spaced instants 1, ..., 5,
has under the natural criterion the edges 1-2, 2-3, 2-4, 3-4 and 4-5 and
unequal degrees,

```{r periodic-degrees}
lengths(natural_visibility_graph(c(0, 1, 0, 1, 0))$adjacency)
```

The engine works by divide and conquer on the maximum. Each level of the
recursion makes at most one decision per point of its intervals, so the number
of decisions is $O(n^2)$ in the worst case (monotone data, whose maximum is
always at an end), $O(n \log n)$ when every maximum splits its interval in
parts of bounded proportion, and $O(n \log n)$ in expectation when the values
are exchangeable and without ties, because the position of each maximum is
then uniform, as the pivot of randomized quicksort. Three facts make the
construction correct for any strictly increasing instants. (i) The maximum $p$ of an interval blocks
every pair across it, because the chord from $a$ to $b$ takes at $t_p$ a
convex combination of $y_a$ and $y_b$, which does not exceed $y_p$. (ii) For
$j < p$, every point between $j$ and $p$ is below the chord from $j$ to $p$ if
and only if the slope from $j$ to $p$ is smaller than the slope from each such
point to $p$; scanning leftwards from $p$, the smallest of those slopes belongs
to the last index found visible, $m$, so $j$ is visible if and only if
$D(j, p, m) > 0$; symmetrically on the right. (iii) The edges inside each half
depend only on that half.

**Exactness.** Each decision is the sign of one quantity $D$, and the engine
computes that sign exactly under the hypotheses on the arithmetic stated in
the help page `?natural_visibility_exactness`. When the six inputs of a decision are at most
$2^{510}$ in magnitude, so that no intermediate value can overflow, it
evaluates in floating point the
orientation determinant $O = -D$, as the difference of two rounded products
$l$ and $r$, and accepts the sign of the result when its magnitude exceeds
$8uS$, with $u = 2^{-53}$ and $S$ the computed value of $|l| + |r|$: above that
threshold the sign is proved correct. Otherwise, near a tie, or when $S$ does
not exceed $2^{-930}$, about $1.1 \times 10^{-280}$, it evaluates $D$
exactly in
integer arithmetic on the binary mantissas and exponents of its six products.
This is the adaptive scheme of Shewchuk (1997). The proofs of both stages,
with their hypotheses on the arithmetic (every IEEE rounding mode; evaluation
in binary64, in a format with at least 64 significand bits, or in the x87
format at 53 bits) and what is checked of them, are written in the help page
`?natural_visibility_exactness`; every intermediate value of the
floating-point evaluation is stored, so that the compiler cannot fuse a
product into the subtraction or reorder the operations. The magnitude limit
is needed: a directed rounding mode can turn an overflow into the largest
finite double instead of an infinity, with a relative error no longer bounded
by $2u$, and a filter without the limit can accept a wrong sign. The graph is therefore the natural
visibility graph of the doubles received, independent of the order in which
floating-point operations round. Both constructors stop with an error when
the floating-point environment does not provide IEEE 754 double arithmetic
(subnormal numbers flushed to zero, or operations with fewer than 53
significant bits), a condition that some libraries create.

**Decimal data.** A decimal such as 0.1 has no exact binary representation,
so points that are collinear in decimal can lie slightly above or below the
line in binary, and the exact criterion follows the binary numbers:

```{r decimal-line}
y <- round(0.1 * (1:40) + 0.3, 1)
natural_visibility_graph(y)$n_edges
```

The 40 points lie on a line in decimal, where only consecutive points see
each other, but their binary values do not. Suppose the values were recorded
with $d$ decimal places, $q_i = m_i / 10^d$ with a non-negative integer
$d \le 22$ and integers $|m_i| \le 2^{50}$, each stored as the double $y_i$
nearest to it. Then `m <- round(10^d * y)` returns the integers $m_i$, and the
graph of the recorded decimals is the graph of `m`, which the engine computes
exactly. The check `all(abs(m) <= 2^50) && all(m / 10^d == y)` passes in that
case, fails whenever some $y_i$ is not the double nearest to a decimal with
$d$ places, and, when it passes, $m_i / 10^d$ is the only decimal with $d$
places whose nearest double is $y_i$:

```{r decimal-recipe}
m <- round(10 * y)
all(abs(m) <= 2^50) && all(m / 10 == y)
natural_visibility_graph(m)$n_edges
```

The proof runs in the rounding to nearest in which R computes. $10^d$ is an
exact double for $d \le 22$, because $5^{22} < 2^{53}$. Since $y_i$ is nearest
to $q_i$, $y_i = q_i (1 + \delta)$ with $|\delta| < u$; the computed product
`10^d * y[i]` equals $m_i (1 + \delta)(1 + \delta')$ with $|\delta'| < u$, so it
differs from $m_i$ by less than $2^{50}(2u + u^2)$, below $1/2$, and
`round()` returns $m_i$; `m[i] / 10^d`, a single correctly rounded division,
returns the double nearest to $q_i$, which is $y_i$. When the check passes,
$y_i$ is the double nearest to $m_i / 10^d$, and no other decimal with $d$
places shares it, because two such decimals differ by at least $10^{-d}$ while
two numbers rounded to the same $y_i$ differ by at most $2u|y_i| < 10^{-d}/2$.
Multiplying every value by $10^d > 0$ multiplies every $D$ by $10^d$, which
keeps its sign, and the integers $m_i$ are exact doubles. Without the bound on
$|m_i|$ the recipe can fail: $2^{53}$, $2^{53} + 1$ and $2^{53} + 2$ are stored
as $2^{53}$, $2^{53}$ and $2^{53} + 2$, and no rounding recovers the middle
value.

## Unequally spaced instants

Both constructors accept `times`, a numeric, `Date` or date-time vector of
finite, strictly increasing instants of the same length as the series; `NULL`
means $1, \ldots, n$. The natural criterion uses the instants as abscissae;
the horizontal one depends only on their order.

```{r instants}
x <- c(0, 2, 4)
natural_visibility_graph(x)$edges
natural_visibility_graph(x, times = c(0, 1, 4))$edges
natural_visibility_graph(x, times = as.Date(c("2024-01-01", "2024-01-02", "2024-01-05")))$edges
```

With equally spaced instants the three points are collinear and the middle
one blocks the first and third; with the instants 0, 1, 4 the middle point lies
above the chord and blocks too; the dates of the third call fall 0, 1 and 4 days
after the first, so they give the same graph as the instants 0, 1, 4. The
reversal of this series, in the section on theorem T1, is where unequal
spacing changes the answer.

# Directed topologies and bitopological spaces

## Forward and backward topologies

In the directed visibility graph every arc goes from the earlier to the later
observation. The *forward topology* $\tau^+$ is generated by the closed forward
neighbourhoods $N^+[v] = \{v\} \cup \{w : v \to w\}$ and the *backward
topology* $\tau^-$ by the closed backward neighbourhoods
$N^-[v] = \{v\} \cup \{w : w \to v\}$. The triple $(V, \tau^+, \tau^-)$ is a
bitopological space, a set with two topologies, in the sense of Kelly (1963).
`generate_bitopology()` builds the graph and the two topologies, together with
the undirected topology and the Alexandrov topology described next.

```{r bitopology}
series <- c(3, 1, 4, 1, 5, 9, 2, 6)
bt <- generate_bitopology(series, graph_type = "hvg")
bt$forward$components
bt$backward$components
```

## The Alexandrov topology of reachability

The Alexandrov topology of a digraph has as open sets the sets closed under
following arcs; the smallest open set containing $v$ is the set of vertices
reachable from $v$, $v$ included. These reachable sets form its minimal base,
which is not closed under intersection in general:

```{r alexandrov-base}
cross <- generate_alexandrov_topology(list(c(3L, 4L), c(3L, 4L), integer(0), integer(0)), 4L)
cross$base
```

For the arcs $1 \to 3$, $1 \to 4$, $2 \to 3$, $2 \to 4$, the first two reachable
sets meet in $\{3, 4\}$, which is open, as the union of $\{3\}$ and $\{4\}$,
but is not a reachable set. The Alexandrov topology is contained in the
forward topology of every digraph, since each reachable set is the union of
the $N^+[w]$ of its points, but not in the undirected topology: on the path
$1 \to 2 \to 3 \to 4$, $\{4\}$ is open in the first and not in the second,
whose smallest open set containing 4 is $\{3, 4\}$. On a visibility graph,
consecutive observations always see each other, reachability is the total
order, and the Alexandrov topology is the same chain $\{i, \ldots, n\}$ for
every series.

## Invariants

`bitopology_invariants()` reports the numbers of components $C^+$ and $C^-$,
their difference $D = C^- - C^+$ (`asymmetry_direction`) and
$I_C = |C^+ - C^-| / \max(C^+, C^-)$ (`irreversibility_components`), the sizes
of the two bases, and their excess over the Alexandrov base.

```{r invariants}
inv <- bt$invariants
inv[c("forward_components", "backward_components", "asymmetry_direction",
      "irreversibility_components", "forward_base_size", "backward_base_size")]
```

### The two base sizes are equal

**Proposition.** *Let $R$ be a relation on a finite non-empty set $V$, with
rows $F(v) = \{w : v R w\}$ and columns $C(v) = \{w : w R v\}$. The non-empty
intersections of non-empty families of rows are as many as the non-empty
intersections of non-empty families of columns.*

*Proof.* For $A \subseteq V$ let $A' = \bigcap_{v \in A} F(v)$ (with
$\emptyset' = V$), and for $B \subseteq V$ let
$B^* = \{v : B \subseteq F(v)\} = \bigcap_{w \in B} C(w)$. Both maps reverse
inclusion, and $A \subseteq A'^*$, $B \subseteq B^{*\prime}$; hence
$A' = A'^{*\prime}$ and $B^* = B^{*\prime *}$, and the two maps are mutually
inverse bijections between the intersections of rows over all $A$ and the
intersections of columns over all $B$, the *intents* and *extents* of formal
concept analysis (Ganter & Wille, 1999). The family in the statement is the
set of intents without $V$, unless some row is $V$, and without $\emptyset$,
which is an intent exactly when no column is $V$; for columns the two
corrections are the same two events with the roles exchanged. $\square$

The closed forward and backward neighbourhoods are the rows and columns of
the reflexive closure of the arc relation, and the engine keeps exactly the
non-empty intersections, so the two base sizes coincide whenever both bases
are complete, and so do the two gains over the Alexandrov base, which counts
the strongly connected components in either direction. The gains are counts
of the construction: they are non-negative, but they do not measure how much
finer $\tau^+$ is than the Alexandrov topology, since the size of an
intersection-closed base depends on the subbase that generates it.

### Time reversal: theorem T1

**Theorem (T1).** *Let a series $x$ be observed at strictly increasing
instants $t$, let $\tilde{x}_i = x_{n+1-i}$, $\tilde{t}_i = -t_{n+1-i}$ and
$r(i) = n + 1 - i$. Under either visibility criterion, $\{i, j\}$ is an edge
for $(\tilde{x}, \tilde{t})$ if and only if $\{r(i), r(j)\}$ is an edge for
$(x, t)$. The map $r$ is a homeomorphism from $\tau^+(\tilde{x}, \tilde{t})$
onto $\tau^-(x, t)$ and from $\tau^-(\tilde{x}, \tilde{t})$ onto
$\tau^+(x, t)$; hence $C^+$ and $C^-$ are exchanged, the base sizes are
exchanged, $D$ changes sign and $I_C$ is unchanged.*

In formulas, $C^+(\tilde{x}, \tilde{t}) = C^-(x, t)$, so
$D = C^- - C^+$ changes sign and $I_C = |C^+ - C^-| / \max(C^+, C^-)$ does
not change.

*Proof.* The horizontal criterion for $i < j$ involves only the values strictly
between them, which $r$ carries onto those strictly between $r(j)$ and $r(i)$.
For the natural one, $D(r(b), r(a), r(k))$ computed for
$(\tilde{x}, \tilde{t})$ equals $D(a, b, k)$ computed for $(x, t)$, term by
term. Arcs run from the lower index to the higher, and $r$ inverts that
order, so closed forward neighbourhoods become closed backward ones; a
bijection between subbases that commutes with the point map is a
homeomorphism of the topologies they generate and carries bases and
components onto each other. $\square$

`time_reverse()` performs the reversal. Reversing the values alone over the
same instants is a different operation as soon as the gaps between instants
are not palindromic, and for the natural criterion it can change the graph:

```{r t1-check}
t <- c(0, 1, 4)
x <- c(0, 2, 4)
rv <- time_reverse(x, t)
rv
natural_visibility_graph(x, times = t)$edges
natural_visibility_graph(rv$series, times = rv$times)$edges
natural_visibility_graph(rev(x), times = t)$edges
```

With the reflected instants the edge sets correspond under $r$; with the
same instants the reversed values make the first and third observations
visible to each other, which the original series does not. Because the
natural criterion is decided exactly under the hypotheses of
`?natural_visibility_exactness` and negation is exact, T1 holds in the
computation for every input:

```{r t1-random}
set.seed(2026)
x <- rnorm(60)
t <- cumsum(rexp(60))
rv <- time_reverse(x, t)
d <- generate_bitopology(x, "nvg", times = t)$invariants$asymmetry_direction
d_rev <- generate_bitopology(rv$series, "nvg", times = rv$times)$invariants$asymmetry_direction
c(d, d_rev)
```

### What the invariants do not say

T1 fixes the meaning of the sign of $D$; it has no converse. When a series and
the gaps between its instants read the same in both directions (with equally
spaced instants, a palindrome), the reversed series is the same series up to a
translation of time, and $D = 0$ and $I_C = 0$ exactly. But a non-zero $D$ does
not show that the process that generated the series is irreversible, a zero
$D$ does not show that it is reversible, and the sign of $D$ is not a
prediction attached to any particular shape of the series. The probabilistic
statement T1 supports concerns laws: if the joint law of the observations
together with their instants is invariant under the joint reversal (values
read backwards, instants reflected), the law of $D$ is symmetric about zero,
the mass at zero included; with equally spaced instants, or with fixed
instants whose gaps read the same in both directions, the condition is that
the law of the values equals the law of their reversal, and so it is for the
horizontal criterion with any fixed instants, which it does not use. For the
natural criterion with fixed instants whose gaps are not palindromic,
reversing the values alone is another operation, to which the statement does
not apply. None of this says anything about stationarity or
about the generating process. Lacasa et al. (2012) treat time irreversibility
statistically with visibility graphs, through the Kullback-Leibler divergence
between the in- and out-degree distributions of the directed horizontal
graph.

### Pairwise connectedness

The bitopological space is *pairwise disconnected* when $V$ is the union of
two disjoint non-empty sets, the first open in $\tau^+$ and the second open in
$\tau^-$. `bitopology_invariants()` decides it exactly from the two subbases,
without enumerating either topology. With $U^+_x$ and $U^-_x$ the smallest
open sets of $x$, let $H$ be the digraph with an arc $x \to y$ when
$y \in U^+_x$ or $x \in U^-_y$. A set $A$ is open in $\tau^+$ with its
complement open in $\tau^-$ exactly when no arc of $H$ leaves $A$, because
$A \in \tau^+$ means $U^+_x \subseteq A$ for $x \in A$, and
$V \setminus A \in \tau^-$ means that $x \in A$ and $x \in U^-_y$ imply
$y \in A$. Such a proper non-empty set exists if and only if $H$ is not
strongly connected, and a strongly connected component that no arc leaves is
one. The accompanying article searches the enumerated topologies instead (its
algorithm 5), with an undecided answer when an enumeration is cut; both give
the same answer whenever the search decides.

On a directed visibility graph the answer is known before computing anything.
Every arc goes from an earlier to a later instant, so the closed
out-neighbourhood of each point lies inside the final segment that starts at
it, and each final segment $\{k+1, \ldots, n\}$, the union of the closed
out-neighbourhoods of its points, is open in $\tau^+$; in the same way each
initial segment $\{1, \ldots, k\}$ is open in $\tau^-$. The two segments
partition $V$, so the space is pairwise disconnected for every series with
$n \ge 2$, and the check carries no information about the series; the
witness it returns is one found by the search, valid but not necessarily the
pair of segments:

```{r pairwise}
small <- generate_bitopology(c(3, 1, 4))
small$invariants$pairwise
```

The check can answer `TRUE` on digraphs with cycles. With the arcs
$1 \to 2$, $2 \to 1$, $2 \to 3$, $3 \to 1$ and $3 \to 2$, the forward
topology is $\{\emptyset, \{1, 2\}, V\}$ and the backward one
$\{\emptyset, \{2, 3\}, V\}$; the complement of $\{1, 2\}$ is not open in the
backward topology, and the space is pairwise connected although the two
topologies differ:

```{r pairwise-cycle}
out_adj <- list(2L, c(1L, 3L), c(1L, 2L))
in_adj <- list(c(2L, 3L), c(1L, 3L), 2L)
tf <- generate_topology(out_adj, 3L, max_open_sets = 100L)
tb <- generate_topology(in_adj, 3L, max_open_sets = 100L)
tf$topology
tb$topology
bitopology_invariants(tf, tb, 3L)$pairwise$pairwise_connected
```

# Implementation notes

Subsets of $\{1, \ldots, n\}$ are stored as bitsets of $W = \lceil n/64 \rceil$
words of 64 bits, so that intersection, union, complement and inclusion take
$O(W)$ word operations. The engines are templates on $W$, instantiated for
$W = 1, 2, 3$ ($n \le 192$), where the compiler unrolls every loop, and a
runtime version covers larger $n$ without limit. Families of sets are
deduplicated in open-addressing hash tables whose hash mixes each word with
the finalizer of the generator known as splitmix64.

| Step | Cost |
|:-----|:-----|
| Horizontal visibility graph | $O(n)$ |
| Natural visibility graph | $O(n^2)$ decisions in the worst case; $O(n \log n)$ when the maxima split the intervals in parts of bounded proportion, and in expectation for exchangeable values without ties; a decision that goes to the exact evaluation costs more, by a bounded factor |
| Base (intersection closure) | $O(b^2 W)$ expected for a base of $b$ sets (hashed deduplication) |
| Components | $O(n^2 \lceil b/64 \rceil)$ |
| Topology (union closure) | $O(|\tau|^2 W)$ expected, with $|\tau| \le 2^n$ |
| Alexandrov reachability, index-ordered digraph | $O(n m / 64)$ for $m$ arcs |
| Pairwise connectedness | $O(\sum_S |S|^2)$ expected over the members $S$ of the two subbases, plus a depth-first search over the arcs of $H$; at most $O(n^3)$ |

The costs marked as expected rest on hash tables. They assume that the hash
spreads the stored sets evenly over the table, the usual uniform-hashing
assumption; a fixed hash function gives no worst-case guarantee, and in the
worst case a lookup takes time proportional to the number of sets stored.

# A worked example

The following series is simulated, so that the example is reproducible; the
numbers it prints describe this series and are not findings about any
process.

```{r worked}
set.seed(11)
s <- cumsum(rnorm(120))
for (type in c("hvg", "nvg")) {
  b <- generate_bitopology(s, graph_type = type)
  cat(type, ": edges", b$graph$n_edges,
      "| C+", b$invariants$forward_components,
      "| C-", b$invariants$backward_components,
      "| D", b$invariants$asymmetry_direction,
      "| base sizes", b$invariants$forward_base_size, b$invariants$backward_base_size,
      "| undirected components", b$invariants$undirected_info$undirected_components, "\n")
}
```

The two base sizes agree, as the proposition of the section on the equality of the two base sizes
requires. Reversing the series changes the sign of $D$ and leaves the rest
exchanged, as T1 requires:

```{r worked-reversed}
r <- time_reverse(s)
generate_bitopology(r$series, graph_type = "nvg")$invariants[c("forward_components",
                                                                "backward_components",
                                                                "asymmetry_direction")]
```

# Legacy threshold functions

Version 0.1.0 built topologies from metric thresholds on the values instead of
graphs, and those functions are kept:

* `calculate_thresholds()` returns five candidate radii: the mean and median
  gap between consecutive sorted values, the standard deviation, a quarter of
  the interquartile range, and an order statistic of all pairwise distances
  named `dbscan` for historical reasons.
* `calculate_topology()` returns the number of sets in the base generated by
  the neighbourhoods $\{j : |x_j - x_i| \le h\}$, counting the empty set and
  the whole index set.
* `analyze_topology_factors()` and `visualize_topology_thresholds()` tabulate
  and plot those numbers for several radii.
* `is_topology_connected()` reports whether the graph of co-occurrence in a
  family of sets is connected, a necessary condition for connectedness when
  the family consists of open sets that cover the space;
  `is_topology_connected2()` is a label-dependent reachability check that is
  neither necessary nor sufficient; `is_topology_connected_manual()` checks
  coverage.
* `simplest_topology()` returns the discrete topology and
  `complete_topology()` the topology of the complete graph, the indiscrete one.

The radius of a threshold construction is a choice of the user; visibility
graphs need none.

# Function index

| Function | Purpose |
|:---------|:--------|
| `horizontal_visibility_graph()` | horizontal visibility graph, optionally directed, with instants |
| `natural_visibility_graph()` | natural visibility graph, exact criterion, with instants |
| `time_reverse()` | a series read backwards with its instants reflected |
| `generate_topology()` | topology of closed neighbourhoods, exact components, enumeration |
| `generate_alexandrov_topology()` | Alexandrov topology of reachability of any digraph |
| `generate_bitopology()` | the whole pipeline for a time series |
| `bitopology_invariants()` | invariants of a forward-backward pair |
| `is_topology_connected_exact()` | connectedness of an enumerated topology |
| `calculate_thresholds()`, `calculate_topology()`, `analyze_topology_factors()`, `visualize_topology_thresholds()` | legacy threshold constructions |
| `is_topology_connected()`, `is_topology_connected2()`, `is_topology_connected_manual()` | legacy diagnostics on families of sets |
| `simplest_topology()`, `complete_topology()` | the discrete and the indiscrete reference topologies |

Each function documents its arguments, return value and design decisions on
its help page, for instance `?generate_bitopology`.

# References

Ganter, B., & Wille, R. (1999). *Formal concept analysis: Mathematical
foundations*. Springer. https://doi.org/10.1007/978-3-642-59830-2

Kelly, J. C. (1963). Bitopological spaces. *Proceedings of the London
Mathematical Society*, *s3-13*(1), 71-89. https://doi.org/10.1112/plms/s3-13.1.71

Lacasa, L., Luque, B., Ballesteros, F., Luque, J., & Nuño, J. C. (2008). From
time series to complex networks: The visibility graph. *Proceedings of the
National Academy of Sciences*, *105*(13), 4972-4975.
https://doi.org/10.1073/pnas.0709247105

Lacasa, L., Nuñez, A., Roldán, É., Parrondo, J. M. R., & Luque, B. (2012). Time
series irreversibility: A visibility graph approach. *The European Physical
Journal B*, *85*(6), Article 217. https://doi.org/10.1140/epjb/e2012-20809-8

Luque, B., Lacasa, L., Ballesteros, F., & Luque, J. (2009). Horizontal
visibility graphs: Exact results for random time series. *Physical Review E*,
*80*(4), Article 046103. https://doi.org/10.1103/PhysRevE.80.046103

Nada, S., El Atik, A. E. F., & Atef, M. (2018). New types of topological
structures via graphs. *Mathematical Methods in the Applied Sciences*,
*41*(15), 5801-5810. https://doi.org/10.1002/mma.4726

Shewchuk, J. R. (1997). Adaptive precision floating-point arithmetic and fast
robust geometric predicates. *Discrete & Computational Geometry*, *18*(3),
305-363. https://doi.org/10.1007/PL00009321
